---
title: "My First Notebook MDA2024"
output: html_notebook
---

# Introduction


# 1. Data Description

# 1.1. Sample Selected

You can define bullet list or numbered list:

- Pizza
- Pasta
- Cafe
  - Espresso
  - Macchiato
  - Cappuccino
- Vodka
  - Bisont Grass
  - Soplica

# 1.2. Formula

Here you can define formula

$$ y = \beta_0 + \beta_1X + \epsilon $$
the formula can be reported in the text: $\mu = 1/n \sum X_i$

# 2. Data Analysis
To add R code in the Notebook we need to use the **Chunk**.

```{r}
X <- iris
```

It is possible to have an overview of the data by using the *summary* function.

```{r}
summary(X)
```
In R there are three main type of data:

- **Matrix**. Mathematical Object. In our example **Y** is a matrix.
- **Data Frame**. It is devoted to organize and analyze data and it is a generalization of the Matrix. In our example, **X** is a data frame.
- **List**. It is an object that can include Data Frames, Matrices or other lists.

```{r}
Y <- as.matrix(X[ ,1:4])
```

To handle data you can use the following code:

```{r}
X[10, 2]       # selection of one element in the Data Frame (or matrix)
X[5:20, 1:3]   # selection of an interval
X[5:20, ]      # the empty space select all the columns or rows
X$Sepal.Length # the symbol $ is used to select a column in the data frame
```
# 2.1. Plots in R

```{r}
boxplot(X$Sepal.Length, main = "Box Plot of the Sepal Length of the IRIS Flowers", col = "blue", horizontal = T)
```
The elements of the box-plot are reported below:

- **Q1**. It is the *First Quartile*. It leaves 25% of units on the left and 75% of units on the right. *Left side of the box*
- **Me**. It is the *Median=Q2*. It leaves 50% of units on left and right. *Bold Line in the middle of the box*
- **Q3**. It is the *Third Quartile*. It leaves 75% of the units on the left and 25% of the units on the right. *Right side of the box*
- In case of **no outliers**, the whiskers are defined as:
  - **Xmin** is the left whiskers
  - **Xmax** is the right whiskers
- In case of **outliers**, the whiskers are defined as:
  - **Linf** = Q1 - 1.5 (Q3-Q1): Lower Limit
  - **Lsup** = Q1 + 1.5 (Q3-Q1): Upper Limit
  
```{r}
# no.1.1
boxplot(X[ ,1:4])
# no.1.2
boxplot(X[ ,1:4], main = "Box Plot of all quantitative variables of IRIS data", col = terrain.colors(4))
```
  



