---
title: "My First Notebook MDA2024"
output: html_notebook
---

# Introduction


# 1. Data Description

# 1.1. Sample Selected

You can define bullet list or numbered list:

- Pizza
- Pasta
- Cafe
  - Espresso
  - Macchiato
  - Cappuccino
- Vodka
  - Bisont Grass
  - Soplica

# 1.2 Formula

Here you can define formula

$$ y = \beta_0 + \beta_1X + \epsilon $$
the formula can be reported in the text: $\mu = 1/n \sum X_i$

# 2. Data Analysis
To add R code in the Notebook we need to use the **Chunk**.

```{r}
X <- iris
```

It is possible to have an overview of the data by using the *summary* function.

```{r}
summary(X)
```
In R there are three main type of data:

- **Matrix**. Mathematical Object. In our example **Y** is a matrix.
- **Data Frame**. It is devoted to organize and analyze data and it is a generalization of the Matrix. In our example, **X** is a data frame.
- **List**. It is an object that can include Data Frames, Matrices or other lists.

```{r}
Y <- as.matrix(X[ ,1:4])
```

To handle data you can use the following code:

```{r}
X[10, 2]     # selection of one element in the Data Frame (or matrix)
X[5:20, 1:3] # selection of an interval
X[5:20, ]    # the empty space select all the columns or rows
```





